---
title: "R tidyverse - VDAB open data example map"
author: "Maarten Hermans - @hermansm"
output: 
  html_notebook: 
    toc: yes
    toc_float: yes
---

# Load required libraries

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(sf)
library(jsonlite)
library(leaflet)
library(stringr)
library(htmlwidgets)

# if needed, run first:
# install.packages("BelgiumMaps.StatBel", repos = "http://www.datatailor.be/rcube", type = "source")
library(BelgiumMaps.StatBel) 
```

# Load/fetch requried data

Used in example:

* [VDAB office-locations as open data](https://www.vdab.be/trends/open_data/locaties) (JSON-file).
* Administrative boundaries for Belgian *arrondissementen*, courtesy of the [BelgiumMaps.StatBel](https://github.com/bnosac/BelgiumMaps.StatBel) R package. 

## Load arrondissement borders

```{r, warning=FALSE, message=FALSE}
# load district/arrondissement spatial structure from package BelgiumMaps.StatBel
data('BE_ADMIN_DISTRICT') 

# convert to simple features dataset-structure
arronds = st_as_sf(BE_ADMIN_DISTRICT) 

# filter on on arrondissementen located in Flanders and Brussels
arronds = arronds %>%
  filter(TX_RGN_DESCR_NL %in% c('Vlaams Gewest', 'Brussels Hoofdstedelijk Gewest'))
```

```{r}
# rough plot, to verify that spatial data is as expected
plot(arronds, max.plot = 1)
```

## Fetch open data VDAB office locations 

```{r}
# fetch open data in JSON-format
vdab.kantoren = as_tibble(fromJSON('http://opendata.vdab.be/vdab/locaties.json'))
vdab.kantoren # dataset with office locations, including coordinates
```


## Describe and change VDAB-data


```{r}
# count te different types of VDAB offices
vdab.kantoren %>%
  group_by(typelocatie) %>%
  tally()
```


```{r}
vdab.kantoren = vdab.kantoren %>%
  mutate(
    # add a variable 'popup' with the the HTML-snippet per office,
    # which  will be shown in the map-popup
    popup = str_glue("<h3>{title}</h3><br /><b>Type: </b>{typelocatie}<br /><b>Adres</b>:  {straatNr} {plaats}<br /><b>Teleloon</b>: {telefoonnummer}"),
    
    # make sure that coordinates are not a character, but numeric variables
    lat = as.numeric(lat),
    lon = as.numeric(lon))
```



# Construct interactive map

## Minimal map example

```{r, message=FALSE}
# basic interactive map with a oneliner
# (width-argument not needed, only for online-output)
leaflet(vdab.kantoren, width = '100%') %>% addTiles() %>% addMarkers()
```

## Styled map example w/t popups

Styling:

* Popups with VDAB-office information.
* Custom tile background
* Custom color & symbol markers, based on discrete values (office-type)
* Overlay of *arrondissement*-polygons: now grey/white-styled, could be colored based on e.g. unemployment-rate. 

```{r}
# create two icons with different color & symbol for the two types of VDAB-offices
# for different symbols, cf. https://rstudio.github.io/leaflet/markers.html 
kantoor_icons <- awesomeIconList(
  'kantoor met onthaal' = makeAwesomeIcon(
    icon = "user", 
    library = "fa", 
    markerColor = "blue"),
  'opleidingscentrum' = makeAwesomeIcon(
    icon = "graduation-cap", 
    library = "fa", 
    markerColor = "red"))
```

```{r}
m.kantoren = leaflet(arronds, width = '100%') %>% 
  # add minimalistic-style map blackground tiles
  addProviderTiles(providers$CartoDB.Positron) %>% 
  
  # add grey arrondissement polygons w/t white border
  addPolygons(fillColor =  'grey20', color = 'white') %>% 
  
  # add custom styled markers w/t popups
  addAwesomeMarkers(
    data = vdab.kantoren,
    lng = ~lon, lat = ~lat,
    icon = ~kantoor_icons[typelocatie],
    popup = ~popup)
```

```{r}
m.kantoren
```

# Save stand-alone map

```{r}
saveWidget(m.kantoren, 'vdab_api_popup_map.html')
```

