---
title: "Using OSM-data and WSM-services for historical data and maps"
output: html_notebook
---


```{r, warning=FALSE, message=FALSE}
library(osmdata)
library(stringr)
library(dplyr)
library(sf)
library(mapview)
library(leaflet)
```

Download all spatial features in a bounding box around Leuven from Open Street Maps, and filter on type 'highway' (streets?).

```{r}
q.leuven <- opq(bbox = 'Leuven, Belgium') %>%
  add_osm_feature(key = 'highway') %>%
  add_osm_feature(key = 'name') %>%
  osmdata_sf()
```

Select all streets in Leuven which have "*gang*" in the streetname. 

```{r, warning=FALSE}
gangen = q.leuven$osm_lines %>%
  select(name) %>%
  filter(str_detect(name, 'gang') == TRUE)
names(gangen$geometry) <- NULL # https://stackoverflow.com/a/50846998/125085
```

View all "gangen" in Leuven. Pop-up works for mapview(), TODO: fix on regular layer.

```{r}
mapview(gangen, color = 'blue')
```

Put the Ferrariskaart (1777) and the Vandermaele kaart (1846-1856) on two selectable layers, and center on Leuven.

```{r}
m.leuven <- leaflet() %>% 
  setView(4.69839084, 50.880089, zoom = 14) %>%
  addWMSTiles(
    group = 'Ferrariskaart (1777)',
    "https://geoservices.informatievlaanderen.be/raadpleegdiensten/histcart/wms",
    layers = list("ferraris"),
      options = WMSTileOptions(format = "image/png", transparent = FALSE),
    attribution = "Source: GIS Geoservices Informatie Vlaanderen") %>%
  addWMSTiles(
    group = 'Kaart Vandermaelen (1846-1854)',
    "https://geoservices.informatievlaanderen.be/raadpleegdiensten/histcart/wms",
    layers = list("vandermaelen"),
      options = WMSTileOptions(format = "image/png", transparent = FALSE),
    attribution = "Source: GIS Geoservices Informatie Vlaanderen") %>%
    addTiles(layerId = 'Kaart Leuven vandaag') %>%
  addLayersControl(baseGroups = c('Kaart Vandermaelen (1846-1854)', 'Ferrariskaart (1777)', 'Kaart Leuven vandaag'))
```

```{r}
m.leuven
```

Visualise the gangen on the base layer.

```{r}
leaflet(gangen) %>%
  addTiles() %>%
  addPolylines()
```

Visualise the gangen on the Ferraris and Vandermaele kaarten.

```{r}
m.leuven %>%
  addPolylines(data = gangen)
```

# TODO: fix pop-ups with information about gangen

```{r}
mapview::addFeatures(map = m.leuven, 
                           data = gangen, 
                           weight = 6, 
                           fillColor = 'purple', 
                           color = 'purple')
```
